### 2018-04-27 response.tex

\comment{7}{%
The two questions posed in the introduction lines 59--61: I don't think the first one is really the question, all parameters should be considered if anyone has thought of them as important (i.e., has proposed them). The second question is the main question and the one being addressed in this paper.
}

\response{%
We thank the reviewer for expressing his/her opinion about this point. First, we changed slightly the sentences as a whole to emphasize these are indirect questions as opposed to direct questions. That said, we do think they both are valid and relevant questions. The argument made by the reviewer that any or all parameters should be included if someone has thought of them as important is precisely one of the points we argue against, and the results of this research support that argument. The fact that someone thinks of a parameter as relevant does not guarantee its relevance, it is only the reflect of his/her expert (though still subjective) opinion. There may be parameters that are more influential, and these can be as influential as to the point of proving others irrelevant. This is the advantage of the procedure we follow here, which allows one to evaluate objectively, based on how data relates to the results, how relevant or irrelevant is each one of the metrics involved in reaching the same conclusions. This, in turn, allows one to prioritize the metrics, which is the point of the second indirect question posed in this part of the introduction.
}

\comment{8}{%
You are treating these metrics as ``independent variables'', but as you have also noted many of them are not. For example, Arias intensity and energy are correlated. Does this compromise your conclusions? If not explain why in the Results section.
}
%
\response{%
This is a very valid point. However, an important aspect to keep in mind is that the clustering process and the subsequent construction of the decision is not the same as a correlation analysis, and it is therefore not compromised by having highly correlated metrics. On the contrary, the process naturally selects the metrics with the best chance of predicting the outcome. In the case of two metrics being highly correlated, the process would still analyze which of the two is a better predictor. In some cases it may decide to keep both, and in some cases it may decide that one is enough. In attention to this and other related comments we decided to include an additional table (Table 7) showing how the level of participation of the different metrics would change if one removes some of the initially most relevant metrics. This should help the readers better understand the results of our work.
}

@@

\comment{11}{%
Page 14, line 323: It's not clear how you ``replicate data'', provide more detail.
}

\response{%
We modified this paragraph to provide more details about the oversampling process.
}

\comment{12}{%
I don't think ``confusion matrix'' given in tables is explained in the paper!
}

\response{%
The explanation about confusion matrices in the original manuscript was in the section titled \textbf{Decision Trees}, page 13, between lines 280 and 282. The explanation included an illustrative example for the simplest case of two categories. We assumed it would be clear that this concept is extendable to larger matrics such as those shown in Tables 2 and 3. We added a sentence in the explanation to make it explicit that they can be extended to deal with more attributes.
}

@@

\comment{15}{%
Figure 8: I don't know what you mean by y-label ``decision attributes''. Do you mean ``the number of GOF metrics''? why there is no data point for 1 or 11?
}

\response{%
That is right. It should be the number of decision attributes. We modified the figure accordingly. The reason why there is no data for 1 is because there are no trees that would classify the data using a single metric into the four classes because these are binary trees, i.e., each node represents a binary decision, so one needs at least 2 metrics with 3 or more nodes with a minimum depth of 2. On the other end of things, the reason why there is no data for 11 metrics is because the analysis of the data resulted in the metric C11 not having any participation in the decision process, as shown in the original Table 6. In new version of Table 6 we include cases in which we forcefully removed metric C8, first, and then C5 and C8. Once one does that, the participations of the metrics changes and so would the topology of the trees.
}

\comment{16}{%
The conclusion basically says that response spectrum is the most important validation metric. This is not surprising to any engineer or modeler. The big question is what other parameters describe the differences between a simulation and a recorded motion if their response spectra matched? Or in other words, what are the other important validation parameters aside from response spectrum (many have argued that's duration, but no statistical proof). Can you use your method to answer this question? Can you condition your machine learning process on Sa being the same (for example use a subset of data with similar Sa or artificially make them equal), then decide what's the next important parameter? Would the result be the same as in your conclusions? Or would it be influenced by removal of Sa as a parameter (this is possible since there are correlations between Sa and other metrics).
}

\response{%
We also mention in the conclusion that the next metrics in line are the energy (C4), the peak acceleration (C5) and peak velocity (C6). In order to emphasize this point, as described in the response to comment \#8, we also included a new table (Table 7) showing the participation of the different metrics in the absence of the response spectra (C8) and the peak acceleration (C5). The results confirm that in such cases the scond most relevant metrics are still the energy (C4) and the peak ground velocity (C6) as mentioned before. This additional analysis also confirms the very low or null participation of the duration (C11) in other trees.
}

@@

\introcomment{~}{%
As a strong ground motion seismologist (i.e., I assume I'm part of the audience you hope to reach), there is a lot of machine learning-related jargon that makes certain sections difficult to follow.
}

\response{%
\textcolor{red}{[ *** Pending ***]}
% We agree with you that the technical terms are not commonly used in seismology and strong motion earthquake engineering. The confusion happened because of integrating three different topics into a method to carry out this research. One technical term in one field is different in another field and that cause a confusion. For example cluster, class, and label all are referring to the same concept, however, it depends on the application. We added a paragraph in the \textbf{Data Analysis Method} section to elaborate this confusion.
}

\introcomment{~}{%
It would also be useful to have some discussion regarding which parameters were found to be repetitive, in case future authors wish to look at more than the four recommended parameters.
}

\response{%
We addressed this by considering additional trees in the cases of removing metrics C5 and C8 (see Table 7). This was also done in response to the comments \#8 and \#16 from Reviewer 1. As we explain in the response to those comments, the procedure used here is different from a simpler correlation analisis in which a metric can be directly related by other if ``repetitive''. Here, the algorithm searches for the potential of each metric to serve as a predictor. In some cases, one may need two well correlated metrics to aid in that decision process, while not in others. By including the results shown in the new Table 7, instead of trying to understand these correlations, we can see what metrics become relevant in the absence of others. The results indicate that in addition to the energy and the response spectra, the next in line are the peak response in acceleration, velocity, and the Arias intensity. They also confirm the lack of relevance of the duration (C11), the time integrals (C1) and (C2), and the cross correlation (C10).
}

@@

\comment{5}{%
Lines 59--61, 391--394: It should be stated somewhere that this is really intended for \textbf{engineering purposes}.
}

\response{%
While it is true that the initial intent of Anderson (2004) was to develop credibility of synthetic seismograms for engineering applications, we do not share the need to make that intent explicit here. On the one hand, these metrics have already been used by several authors who did not intend to validate seismograms for any engineering application, but to simply show---by means of them---that their synthetics were realistic in a general sense. On the other hand, we do not want to convey the idea that the steps that we followed are exclusive to narrowing validation criteria for engineering applications. Others may see our work and these steps as applicable to other ends. That being said, we hope the reviewer will be pleased to notice that we modified the Introduction to point out that engineering applications are the one of the motivating factors for the increased attention in verification and validation. Please see the first sentence in the introduction (see line 29).}

\comment{6}{%
Line 61: This sentence should end with a ``?''.
}

\response{%
In our view, these sentences are written as indirect questions and do not require the use of a question mark. To make this clearer, we modified the beginning of the sentences slightly by removing the colon and adding the word ``regarding''.
}

### 2018-04-27 dataset.tex
updated

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/pdf/figure-02}
    \caption{Validation results obtained by \citet{Taborda_2014_BSSA} in the form of GOF values across the region of interest obtained from comparisons between simulations for three different velocity models and data recorded at ground motion monitoring stations. The color version of this figure is available only in the electronic edition.}
    \label{fig:ref-gof-maps}
\end{figure*}
% 
\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.48\textwidth]{figures/pdf/figure-03}
    \caption{Statistical distribution of the GOF dataset obtained from \citet{Taborda_2014_BSSA} shown in the form of box-plots for each metric (C1 through C11, see Table \ref{tab:metrics}), velocity model (CVM-S, CVM-H, and CVM-H+GTL), and component\myrevision{s} of motion (NS, EW, UD). In each case, \change{the boxes represent the interquartile range ($\mathrm{IQR} = \mathrm{Q}3 - \mathrm{Q}1$), the medians are indicated by a notch in the boxes, and the vertical lines show} the range of the data, with outliers (data less than $\mathrm{Q}1 - 1.5 \mathrm{IQR}$ and greater than $\mathrm{Q}3 + 1.5 \mathrm{IQR}$) shown as scattered dots. The color version of this figure is available only in the electronic edition.}
    \label{fig:data-box-plot}
\end{figure}

>>>>

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/pdf/figure-02}
    \caption{Validation results obtained by \citet{Taborda_2014_BSSA} in the form of GOF values across the region of interest obtained from comparisons between simulations for three different velocity models and data recorded at ground motion monitoring stations. The color version of this figure is available only in the electronic edition.}
    \label{fig:ref-gof-maps}
\end{figure*}
% 
\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.48\textwidth]{figures/pdf/figure-03}
    \caption{Statistical distribution of the GOF dataset obtained from \citet{Taborda_2014_BSSA} shown in the form of box-plots for each metric (C1 through C11, see Table \ref{tab:metrics}), velocity model (CVM-S, CVM-H, and CVM-H+GTL), and \change{components of motion} (NS, EW, UD). In each case, \change{the boxes represent the interquartile range ($\mathrm{IQR} = \mathrm{Q}3 - \mathrm{Q}1$), the medians are indicated by a notch in the boxes, and the vertical lines show} the range of the data, with outliers (data less than $\mathrm{Q}1 - 1.5 \mathrm{IQR}$ and greater than $\mathrm{Q}3 + 1.5 \mathrm{IQR}$) shown as scattered dots. The color version of this figure is available only in the electronic edition.}
    \label{fig:data-box-plot}
\end{figure}

### 2018-04-27 method.tex
updated

The following sections explain the data processing analysis we put in place to cluster the dataset and obtain \myremove{two} \myrevision{three} alternative decision trees.

>>>>

The following sections explain the data processing analysis we put in place to cluster the dataset and obtain \change{three} alternative decision trees.

### 2018-04-27 results.tex
switching to office computer
updated

Figure \ref{fig:boxed-clusters} shows the statistical distribution (box-plots) of the samples once the dataset is partitioned into the four GOF validation classes. This is similar to Fig.~\ref{fig:data-box-plot}, but after the clustering process is completed. Separately, we prepared similar plots to look at the influence of the velocity models and components of motions on the results of the clustering process and, as observed before, they had no significant differences with the aggregate of all the samples in the dataset. \change{Figure 6 is particularly important because it highlights the outcome of the clustering process and provides a preview into the decision tree analysis results. Notice that metrics C5 through C8 consistently fall within the poor, fair, good, and excellent classification and that metrics C3 and C4 also show a progressive variation in sync with the validation classes. This means that these metrics are likely the best predictors for the outcome of the validation process, as we will see upon performing the decision tree analysis. On the other hand, metrics C1 and C2, and C9 through C11 are almost invariant, therefore less or not decisive in the validation process. C11, in particular, shows a broader (larger boxes) distribution, which indicates that it is less effective in predicting the final validation class.}

@@

Based on these criteria, we selected three candidate trees: T1, T2, and T3. These trees are shown in Fig.~\ref{fig:trees}. T1 is the simplest of the three, and T2 and T3 share some of their topology on the right-hand side. T3 is the most complex of them. More complex trees tend to be deeper, have more nodes, and employ more attributes. \myrevision{complexity of trees is controlled by pruning process.} This leads to higher $F_1$ values, but they may not necessarily be more practical. In total T1, T2, and T3 use 2, 3 and 4 attributes respectively. All coincide in the use of the total energy (C4) and response spectra (C8) as key metrics. T2 adds peak acceleration (C5), and T3 adds peak velocity \change{(C6)} to the previous metrics.

@@

@@

At the bottom of each tree in Fig.~\ref{fig:trees} are the histograms of the data that land on each leaf node. (Note that the count of samples here is done based on the training dataset.) The decision tree assigns the final validation category based on the dominant class in each leaf. As such, the samples in the second leaf-node from the right in T3 are categorized as good despite a significant portion of them being excellent. This is a natural trade-off embedded in the use of \myrevision{pruned} decision trees which actually results in more effective trees.

The actual effectiveness (i.e., the average $F_1$ values shown in Fig.~\ref{fig:nodes-depth}), however, is measured based on the testing dataset. Recall that $F_1$ depends on the number of true predictions and false predictions measured by the confusion matrix and the precision and recall factors. Tables \ref{tab:t1:confusion} and \ref{tab:t3:confusion} show the confusion matrices for T1 and T3, respectively; and Tables \ref{tab:t1:f1} and \ref{tab:t3:f1} show the corresponding results for $P$, $R$, and $F_1$. As it can be seen in the Tables \ref{tab:t1:confusion} and \ref{tab:t3:confusion}, both trees lead to strong diagonal confusion matrices, meaning that the classification into poor, fair, good and excellent is well defined. The differences between both matrices are actually minor and small with respect to their diagonal values, and the $F_1$ values in Tables \ref{tab:t1:f1} and \ref{tab:t3:f1} are all near to or above 0.9. The values obtained for tree T2 are consistent with this.

Another aspect of interest is the level of participation of each metric in the analysis of data as GOF values are run through the decision nodes in the trees. The results of such level of participation are listed in Table \ref{tab:participation} for trees T1, T2, T3, and T66. (The tree T66 is the most complex of them all as inferred from the number of metrics involved and the number of nodes and levels as seen in Fig.~\ref{fig:nodes-depth}.) The percentages in this table represent the amount of data that is seen by a decision node associated by any particular metric. The nodes are not unique and metrics are often used in different nodes in a given tree, thus these percentage are cumulative for each individual metric. A percentage of 100 means a given metric has the opportunity to see all data samples through a single nodes or multiple ones. A low percentage means only a few data samples are evaluated by that metric, and a null percentage means the metric plays no role in the decision tree (it is not present in any node).

Notice that in all trees \myrevision{,using all metrics,} the metrics C4 (total energy) and C8 (response spectra) are consistently the most relevant ones in the decision algorithms. They are followed by C5 (peak acceleration) and C6 (peak velocity). On the other end, notice that C11 (strong phase duration) plays no role whatsoever in any of the trees, and that C1 (Arias integral), C2 (energy integral), and C3 (Arias intensity) have only small to marginal participations. The remaining metrics---C7 (peak displacement), C9 (Fourier spectrum), and C10 (Cross correlation)---have a somewhat significant participation, but only in the more complex trees.

\myrevision{It is obvious that C8(response spectra) and C5(peak acceleration) are the most important metrics in developing the prediction models. Therefore, in order to understand other parameters importance we developed two sets of trees according to complexity of the original trees (the structure of generated trees are not presented here), while once dropping C8 (response spectra) and once dropping C8 (response spectra) and C5(peak acceleration) from the prediction attributes. According to Table \ref{tab:participation}, these trees also emphasize the importance of C4 and C6. However, in the absence of C5(peak acceleration) and C8 (response spectra),  C3 (Arias intensity) and C9 (Fourier spectrum) become more effective in developing prediction models. Also C11(strong phase duration) becomes somewhat important in the most complicated tree.}

A low participation does not necessarily mean that a given metric is not important. Arguably, the strong phase duration is highly regarded as an important parameter for strong motion records in engineering. What the results we present here mean \change{is that, in the context of this particular set of 11 metrics, in order to classify whether a simulation result is poor, fair, good, or excellent, other metrics such as energy (C8) are much more influential in the final result than the duration (C11). This, of course, is done under the assumption that one seeks to narrow the selection of metrics without loss of insight about the outcome of the validation process when compared to the outcome obtained with the whole suite of the eleven metrics used here.}

@@

In the end, the final selection of a preferred tree comes down to reducing complexity in the analysis without compromising the interpretation of results. We favor tree T1 because it is based only in three decision steps, and two GOF metrics: C4 (total energy), and C8 (response spectra).

>>>>

Figure \ref{fig:boxed-clusters} shows the statistical distribution (box-plots) of the samples once the dataset is partitioned into the four GOF validation classes. This is similar to Fig.~\ref{fig:data-box-plot}, but after the clustering process is completed. Separately, we prepared similar plots to look at the influence of the velocity models and components of motions on the results of the clustering process and, as observed before, they had no significant differences with the aggregate of all the samples in the dataset. \change{Figure \ref{fig:boxed-clusters} is particularly important because it highlights the outcome of the clustering process and provides a preview into the decision tree analysis results. Notice that metrics C5 through C8 consistently fall within the poor, fair, good, and excellent classification and that metrics C3 and C4 also show a progressive variation in sync with the validation classes. This means that these metrics are likely the best predictors for the outcome of the validation process, as we will see upon performing the decision tree analysis. On the other hand, metrics C1 and C2, and C9 through C11 are almost invariant, therefore less or not decisive in the validation process. C11, in particular, shows a broader (larger boxes) distribution, which indicates that it is less effective in predicting the final validation class.}

@@

Based on these criteria, we selected three candidate trees: T1, T2, and T3. These trees are shown in Fig.~\ref{fig:trees}. T1 is the simplest of the three, and T2 and T3 share some of their topology on the right-hand side. T3 is the most complex of them. More complex trees tend to be deeper, have more nodes, and employ more attributes\change{, all of which depends---in part---on the pruning process. More complex trees have} higher $F_1$ values, but they may not necessarily be more practical. In total T1, T2, and T3 use 2, 3 and 4 attributes respectively. All coincide in the use of the total energy (C4) and response spectra (C8) as key metrics. T2 adds peak acceleration (C5), and T3 adds peak velocity \change{(C6)} to the previous metrics.

At the bottom of each tree in Fig.~\ref{fig:trees} are the histograms of the data that land on each leaf node. (Note that the count of samples here is done based on the training dataset.) The decision tree assigns the final validation category based on the dominant class in each leaf. As such, the samples in the second leaf-node from the right in T3 are categorized as good despite a significant portion of them being excellent. This is a natural trade-off embedded in the use of decision trees.

% , which actually results in more effective trees.

@@

The actual effectiveness (i.e., the average $F_1$ values shown in Fig.~\ref{fig:nodes-depth}), however, is measured based on the testing dataset. Recall that $F_1$ depends on the number of true predictions and false predictions measured by the confusion matrix and the precision and recall factors. Tables \ref{tab:t1:confusion} and \ref{tab:t3:confusion} show the confusion matrices for T1 and T3, respectively; and Tables \ref{tab:t1:f1} and \ref{tab:t3:f1} show the corresponding results for $P$, $R$, and $F_1$. As it can be seen in the Tables \ref{tab:t1:confusion} and \ref{tab:t3:confusion}, both trees lead to strong diagonal confusion matrices, meaning that the classification into poor, fair, good and excellent is well defined. The differences between both matrices are actually minor and small with respect to their diagonal values, and the $F_1$ values in Tables \ref{tab:t1:f1} and \ref{tab:t3:f1} are all near to or above 0.9. The values obtained for tree T2 are consistent with \change{these observations}.

Another aspect of interest is the level of participation of each metric in the analysis of data, as the GOF values are run through \change{the nodes of the decision trees (i.e., as the tree is traversed)}. The results of such level of participation are listed in Table \ref{tab:participation} for trees T1, T2, T3, and T66. (The tree T66 is the most complex of them all as inferred from the number of metrics involved and the number of nodes and levels seen in Fig.~\ref{fig:nodes-depth}.) The percentages in this table represent the amount of data that is seen by a decision node associated to any particular metric. The nodes are not unique, and metrics are often used in different nodes in a given tree. Therefore, these percentage accumulate for each \change{metric as the tree is traversed}. A percentage of 100 means that a given metric has the opportunity to see all the data samples through one or multiple nodes, a low percentage means only a few data samples are evaluated by that metric, and a null percentage means the metric plays no role in the decision tree (it is not present in any node).

@@

\change{The results in Table 6 highlight the fact that, for all trees,} the metrics C4 (total energy) and C8 (response spectra) are consistently the most relevant ones in the decision algorithms. They are followed by C5 (peak acceleration) and C6 (peak velocity). On the other end, C11 (strong phase duration) plays no role whatsoever in any of the trees, and that C1 (Arias integral), C2 (energy integral), and C3 (Arias intensity) have only small to marginal participations. The remaining metrics---C7 (peak displacement), C9 (Fourier spectrum), and C10 (Cross correlation)---have a somewhat significant participation, but only in the more complex trees.

\change{An additional aspect worth highlighting here is the fact that the clustering and decision tree processes address the problem of correlation between the metrics in a natural way. For instance, one would expect C3 and C5, or C3 and C4 to be related. Considering them all without distinction could lead to double weighing their influence, and assuming a 1-to-1 relationship may overstate their level of correlation. Instead, the process used here determines which of them has a better chance to predict the outcome in the presence of the other, and preserves that metric as part of the decision tree.}

\change{One can put this to test by removing one or more metrics and see how the participation of the metrics rearranges. Table \ref{tab:participation-extra} shows these participations for trees with similar topologies (i) when removing C8, and (ii) when removing C5 and C8. In the first case, C5 takes the place of the most determinant metric, followed by C4 and C6, depending on the tree. In the second case, C6 becomes the most dominant, followed by C3 and C4, which vary in their level of participation depending on the tree. These additional alternatives emphasizes the C4 and C6, and show that other metrics such as C5, C7 and C9 are also relevant. They also reaffirm the low or null levels of contribution of metrics C1, C2, C10 and C11. All this is consistent with what we observed in Fig.~\ref{fig:boxed-clusters}.}

% \myrevision{It is obvious that C8(response spectra) and C5(peak acceleration) are the most important metrics in developing the prediction models. Therefore, in order to understand other parameters importance we developed two sets of trees according to complexity of the original trees (the structure of generated trees are not presented here), while once dropping C8 (response spectra) and once dropping C8 (response spectra) and C5(peak acceleration) from the prediction attributes. According to Table \ref{tab:participation}, these trees also emphasize the importance of C4 and C6. However, in the absence of C5(peak acceleration) and C8 (response spectra),  C3 (Arias intensity) and C9 (Fourier spectrum) become more effective in developing prediction models. Also C11(strong phase duration) becomes somewhat important in the most complicated tree.}

\change{That being said, a} low participation does not necessarily mean that a given metric is not relevant at all. Arguably, the strong phase duration is highly regarded as an important parameter for strong motion records in engineering. What the results we present here mean \change{is that, in the context of this particular set of 11 metrics, in order to classify whether a simulation result is poor, fair, good, or excellent, other metrics such as energy (C8) are much more influential in the final result than the duration (C11). This, of course, is done under the assumption that one seeks to narrow the selection of metrics without loss of insight about the outcome of the validation process when compared to the outcome obtained with the whole suite of the eleven metrics used here.}

@@

In the end, the final selection of a preferred tree comes down to reducing complexity in the analysis without compromising the interpretation of results. We favor tree T1 because it is based only on three decision steps, and two GOF metrics: C4 (total energy), and C8 (response spectra).

### 2018-04-27 testing.tex
updated

\myrevision{It is worth mentioning that, in this study, we presented an integration of scientists knowledge and field data using machine learning algorithms to develop an acceptable and robust prediction model. This method can be used in any field where there are some field data as well as expertise knowledge about the data. Validation metrics are note limited to \citeauthor{Anderson_2004_Proc}'s metrics. Therefore, adding more metrics from different field can be used through the proposed method to develop prediction models. }

>>>>

% \myrevision{It is worth mentioning that, in this study, we presented an integration of scientists knowledge and field data using machine learning algorithms to develop an acceptable and robust prediction model. This method can be used in any field where there are some field data as well as expertise knowledge about the data. Validation metrics are note limited to \citeauthor{Anderson_2004_Proc}'s metrics. Therefore, adding more metrics from different field can be used through the proposed method to develop prediction models. }
